For Finitary Induction-Induction, \\ Induction is Enough

### 2020-05-03 paper/paper.tex
stuff

\begin{definition}[Signature algebra, $\SignAlg$]\label{def:algebra}
  In a model of ETT, a signature algebra is an iterated $\Sigma$ type
  consisting of the following four (families of) sets, 17 operations
  and 18 equalities.
\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\
  & \Con && : \Set \\
  & \Ty  && : \Con\ra\Set \\
  & \Sub  && : \Con\ra\Con\ra\Set \\
  & \Tm  && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Set \\
  & \id && : \Sub\,\Gamma\,\Gamma \\
  & \blank\circ\blank && : \Sub\,\Theta\,\Delta\ra\Sub\,\Gamma\,\Theta\ra\Sub\,\Gamma\,\Delta \\
  & \ass && : (\sigma \circ \delta) \circ \nu = \sigma \circ (\delta \circ \nu) \\
  & \idl && : \id\circ\sigma = \sigma \\
  & \idr && : \sigma\circ\id = \sigma \\
  & \blank[\blank] && : \Ty\,\Delta\ra\Sub\,\Gamma\,\Delta\ra\Ty\,\Gamma \\
  & \blank[\blank] && : \Tm\,\Delta\,A\ra(\sigma:\Sub\,\Gamma\,\Delta)\ra\Tm\,\Gamma\,(A[\sigma]) \\
  & [\id] && : A[\id] = A \\
  & [\circ] && : A[\sigma \circ \delta] = A[\sigma][\delta] \\
  & [\id] && : t[\id] = t \\
  & [\circ] && : t[\sigma \circ \delta] = t[\sigma][\delta] \\
  & \boldsymbol{\cdot} && : \Con \\
  & \epsilon && : \Sub\,\Gamma\,\boldsymbol{\cdot} \\
  & {\boldsymbol{\cdot}\eta} && : (\sigma : \Sub\,\Gamma\,\boldsymbol{\cdot}) \ra \sigma = \epsilon \\
  & \blank\ext\blank && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Con \\
  & \blank,\blank && : (\sigma:\Sub\,\Gamma\,\Delta)\ra\Tm\,\Gamma\,(A[\sigma])\ra\Sub\,\Gamma\,(\Delta\ext A) \\
  & \pi_1 && : \Sub\,\Gamma\,(\Delta\ext A)\ra \Sub\,\Gamma\,\Delta \\
  & \pi_2 && : (\sigma : \Sub\,\Gamma\,(\Delta\ext A))\ra \Tm\,\Gamma\,(A[\pi_1 \sigma]) \\
  & {\pi_1\beta} && : \pi_1(\sigma, t) = \sigma \\
  & {\pi_2\beta} && : \pi_2(\sigma, t) = t \\
  & {\pi\eta} && : (\pi_1\,\sigma, \pi_2\,\sigma) = \sigma \\
  & {,\circ} && : (\sigma, t)\circ\delta = (\sigma\circ\delta, t[\delta]) \\
  & \rlap{(2) Universe} \\
  & \U && : \Ty\,\Gamma \\
  & \El && : \Tm\,\Gamma\,\U\ra\Ty\,\Gamma \\
  & {\U[]} && : \U[\sigma] = \U \\
  & {\El[]} && : (\El\,a)[\sigma] = \El\,(a[\sigma]) \\
  & \rlap{(3) Inductive parameters} \\
  & \Pi && : (a:\Tm\,\Gamma\,\U)\ra\Ty\,(\Gamma\ext\El\,a)\ra\Ty\,\Gamma \\
  & \blank\oldapp\blank && : \Tm\,\Gamma\,(\Pi\,a\,B)\ra (u : \Tm\,\Gamma\,(\El\,a))
  \ra \Tm\,\Gamma\,(\El\,(B[\id,\,u])) \\
  & {\Pi[]} && : (\Pi\,a\,B)[\sigma] = \Pi\,(a[\sigma])\,(B[\sigma\circ\p,\q]) \\
  & {\oldapp[]} && : (t\oldapp\alpha)[\sigma] = (t[\sigma])\mathop{\oldapp}(\alpha[\sigma]) \\
  & \rlap{(4) External parameters} \\
  & \Pim && : (T:\Set)\ra(T\ra\Ty\,\Gamma)\ra\Ty\,\Gamma \\
  & \blank\appm \blank && : \Tm\,\Gamma\,(\Pim\,T\,B)\ra(\alpha:T) \ra\Tm\,\Gamma\,(B\,\alpha) \\
  & {\Pim[]} && : (\Pim\,T\,B)[\sigma] = \Pim\,T\,(\lambda \alpha.(B\,\alpha)[\sigma]) \\
  & {\appm[]} && : (t\appm\alpha)[\sigma] = (t[\sigma])\mathop{\appm}\alpha
\end{alignat*}
Given an $M:\SignAlg$, we denote its components as $\Con^M$, $\Ty^M$,
$\Sub^M$, $\Tm^M$, $\id^M$, and so on. We omit the indices if there is
only one signature algebra in scope (e.g.\ in
\Cref{def:abbrevs,ex:contexts}).
\end{definition}

@@

We define IIT signatures using the Church encoding introduced by
Awodey, Frey and Speight \cite{DBLP:conf/lics/AwodeyFS18}. A
difference is that we avoid the usage of impredicativity as we don't
require that signatures form a signature algebra. It is also not a
problem for our use case that signatures are not in the lowest
universe.
\begin{definition}[IIT signature]
  An IIT signature is a context in an arbitrary signature algebra
  which is compatible with morphisms:
  \begin{alignat*}{6}
    & \Sign :=\,\, && \big({sig}:(M:\SignAlg)\ra\Con^M\big)\times \\
    & && \big((M\,N:\SignAlg)(f:\SignMor\,M\,N)\ra f_\Con\,({sig}\,M) = {sig}\,N \big).
  \end{alignat*}
\end{definition}
Note that as $\Sign$ refers to $\SignAlg$, it lives in a higher
universe level than $\SignAlg$. The compatibility condition says that
if we obtain an $M$-context using ${sig}$ in model $M$ and then we
transport it to $N$ using $f$, we get the same $N$-context as directly
applying ${sig}$ to $N$.

@@

In the previous section, using the notions of signature algebras and
signature morphisms, we defined IIT signatures and what it means for a
model of ETT to support all IITs. In this section we show that if a
model of ETT supports the theory of IIT signatures, then it supports
all IITs.

@@

\begin{definition}[Syntactic signatures]\label{def:syntacticsignature}
  In a model of ETT supporting the theory of ITT signatures, elements
  of $\Con^\I$ are called syntactic signatures. We can turn a
  $({sig},{sigc}) : \Sign$ into $\Con^\I$ by ${sig}\,\I$ and an
  $\Omega:\Con^\I$ into a $\Sign$ by
  $(\lambda M.\ll\Omega\rr_M, \lambda M\,N\,f.f\,(\ll\Omega\rr_M) =
  \ll\Omega\rr_N)$ where the equality in the second component comes
  from uniqueness of the recursor.
\end{definition}

>>>>

\begin{definition}[Signature algebra, $\SignAlg$]\label{def:algebra}
  In a model of ETT, a signature algebra is an iterated $\Sigma$ type
  consisting of the following four (families of) sets, 17 operations
  and 18 equalities.
\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\
  & \Con && : \Set \\
  & \Ty  && : \Con\ra\Set \\
  & \Sub  && : \Con\ra\Con\ra\Set \\
  & \Tm  && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Set \\
  & \id && : \Sub\,\Gamma\,\Gamma \\
  & \blank\circ\blank && : \Sub\,\Theta\,\Delta\ra\Sub\,\Gamma\,\Theta\ra\Sub\,\Gamma\,\Delta \\
  & \ass && : (\sigma \circ \delta) \circ \nu = \sigma \circ (\delta \circ \nu) \\
  & \idl && : \id\circ\sigma = \sigma \\
  & \idr && : \sigma\circ\id = \sigma \\
  & \blank[\blank] && : \Ty\,\Delta\ra\Sub\,\Gamma\,\Delta\ra\Ty\,\Gamma \\
  & \blank[\blank] && : \Tm\,\Delta\,A\ra(\sigma:\Sub\,\Gamma\,\Delta)\ra\Tm\,\Gamma\,(A[\sigma]) \\
  & [\id] && : A[\id] = A \\
  & [\circ] && : A[\sigma \circ \delta] = A[\sigma][\delta] \\
  & [\id] && : t[\id] = t \\
  & [\circ] && : t[\sigma \circ \delta] = t[\sigma][\delta] \\
  & \boldsymbol{\cdot} && : \Con \\
  & \epsilon && : \Sub\,\Gamma\,\boldsymbol{\cdot} \\
  & {\boldsymbol{\cdot}\eta} && : (\sigma : \Sub\,\Gamma\,\boldsymbol{\cdot}) \ra \sigma = \epsilon \\
  & \blank\ext\blank && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Con \\
  & \blank,\blank && : (\sigma:\Sub\,\Gamma\,\Delta)\ra\Tm\,\Gamma\,(A[\sigma])\ra\Sub\,\Gamma\,(\Delta\ext A) \\
  & \pi_1 && : \Sub\,\Gamma\,(\Delta\ext A)\ra \Sub\,\Gamma\,\Delta \\
  & \pi_2 && : (\sigma : \Sub\,\Gamma\,(\Delta\ext A))\ra \Tm\,\Gamma\,(A[\pi_1 \sigma]) \\
  & {\pi_1\beta} && : \pi_1(\sigma, t) = \sigma \\
  & {\pi_2\beta} && : \pi_2(\sigma, t) = t \\
  & {\pi\eta} && : (\pi_1\,\sigma, \pi_2\,\sigma) = \sigma \\
  & {,\circ} && : (\sigma, t)\circ\delta = (\sigma\circ\delta, t[\delta]) \\
  & \rlap{(2) Universe} \\
  & \U && : \Ty\,\Gamma \\
  & \El && : \Tm\,\Gamma\,\U\ra\Ty\,\Gamma \\
  & {\U[]} && : \U[\sigma] = \U \\
  & {\El[]} && : (\El\,a)[\sigma] = \El\,(a[\sigma]) \\
  & \rlap{(3) Inductive parameters} \\
  & \Pi && : (a:\Tm\,\Gamma\,\U)\ra\Ty\,(\Gamma\ext\El\,a)\ra\Ty\,\Gamma \\
  & \blank\oldapp\blank && : \Tm\,\Gamma\,(\Pi\,a\,B)\ra (u : \Tm\,\Gamma\,(\El\,a))
  \ra \Tm\,\Gamma\,(\El\,(B[\id,\,u])) \\
  & {\Pi[]} && : (\Pi\,a\,B)[\sigma] = \Pi\,(a[\sigma])\,(B[\sigma\circ\p,\q]) \\
  & {\oldapp[]} && : (t\oldapp\alpha)[\sigma] = (t[\sigma])\mathop{\oldapp}(\alpha[\sigma]) \\
  & \rlap{(4) External parameters} \\
  & \Pim && : (T:\Set)\ra(T\ra\Ty\,\Gamma)\ra\Ty\,\Gamma \\
  & \blank\appm \blank && : \Tm\,\Gamma\,(\Pim\,T\,B)\ra(\alpha:T) \ra\Tm\,\Gamma\,(B\,\alpha) \\
  & {\Pim[]} && : (\Pim\,T\,B)[\sigma] = \Pim\,T\,(\lambda \alpha.(B\,\alpha)[\sigma]) \\
  & {\appm[]} && : (t\appm\alpha)[\sigma] = (t[\sigma])\mathop{\appm}\alpha
\end{alignat*}
Given an $M:\SignAlg$, we denote its components by $\Con^M$, $\Ty^M$,
$\Sub^M$, $\Tm^M$, $\id^M$, and so on. We omit the indices if there is
only one signature algebra in scope (e.g.\ in
\Cref{def:abbrevs,ex:contexts}).
\end{definition}

@@

We define IIT signatures using the Church encoding introduced by
Awodey, Frey and Speight \cite{DBLP:conf/lics/AwodeyFS18}. A
difference is that we avoid the usage of impredicativity because it is
not a problem for our use case that signatures are not in the lowest
universe.
\begin{definition}[IIT signature]\label{defn:sign}
  An IIT signature is a context in an arbitrary signature algebra that
  is compatible with morphisms:
  \begin{alignat*}{6}
    & \Sign :=\,\, && \big({sig}:(M:\SignAlg)\ra\Con^M\big)\times \\
    & && \big((M\,N:\SignAlg)(f:\SignMor\,M\,N)\ra f_\Con\,({sig}\,M) = {sig}\,N \big).
  \end{alignat*}
\end{definition}
The compatibility condition says that if we obtain an $M$-context
using ${sig}$ in model $M$ and then we transport it to $N$ using $f$,
we get the same $N$-context as directly applying ${sig}$ to $N$.

Our notion of signatures don't form a signature algebra.
\begin{lemma}
  There is no $M : \SignAlg$, in which $\Con^M = \Sign$.
\end{lemma}
\begin{proof}
  If the $\Con$ component in $\SignAlg$ is $\Set_i$, then $\SignAlg$
  is in $\Set_{i+1}$, but as $\Sign$ is defined as
  $(\SignAlg\ra\dots)\times\dots$, it is at least in $\Set_{i+1}$, so
  we can't choose $\Con^M : \Set_i$ to be $\Sign : \Set_{i+1}$.
\end{proof}

@@

In the previous section, using the notions of signature algebras and
signature morphisms, we defined IIT signatures and what it means for a
model of ETT to support all IITs. In this section we show that if a
model of ETT supports the theory of IIT signatures, then it supports
all IITs. In the sense of the Church encoding of \Cref{defn:sign},
every model of ETT can describe ITT signatures. In contrast, in
\Cref{def:theoryofsignatures}, we will require existence of an initial
signature algebra.

@@

\begin{definition}[Syntactic signatures]\label{def:syntacticsignature}
  In a model of ETT supporting the theory of ITT signatures, we call
  elements of $\Con^\I$ syntactic signatures.
\end{definition}

\begin{lemma}
  In a model of ETT supporting the theory of ITT signatures,
  signatures and syntactic signatures are isomorphic.
\end{lemma}
\begin{proof}
  We can turn a $({sig},{sigc}) : \Sign$ into $\Con^\I$ by ${sig}\,\I$
  and an $\Omega:\Con^\I$ into a $\Sign$ by
  $(\lambda M.\ll\Omega\rr_M, \lambda M\,N\,f.f\,(\ll\Omega\rr_M) =
  (f\circ\ll\rr_M)\,\Omega = \ll\Omega\rr_N)$ where the equality in
  the second component comes from uniqueness of the recursor (we have
  to define composition of morphisms $\circ$ for this). The
  compositions of these two maps are the identities: $({sig},{sigc})$
  is mapped to
  $(\lambda M.\ll{sig}\,I\rr_M,\dots) = (\lambda
  M.\ll\blank\rr_M\,({sig}\,I),\dots)$ which is equal to
  $(\lambda M.{sig}\,M,\dots)$ because of ${sigc}$; $\Omega$ is mapped
  to $\ll\Omega\rr_\I = \Omega$ by uniqueness of $\ll\blank\rr$.
\end{proof}
